---
title: "Comparing ccRCC Tumor to Normal Kidney in Public Datasets"
author: "Katja Lamia"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)

```

## Is the expression of bHLH-PAS family members altered in ccRCC?
To address this question, we will analyze publicly available RNA-sequencing data from normal and tumor samples in the TCGA KIRC and the CPTAC datasets, downloaded from the GDC data portal (<https://portal.gdc.cancer.gov/>). 

Note: Only ARNT was detected by mass spec in the CPTAC-3 project (See Clark et al., 2019 *Cell*, <https://pubmed.ncbi.nlm.nih.gov/31675502/>). ARNT protein was observed to be increased in ccRCC tumors compared to normal kidney.

```{r load file paths and categorize into groups, include=FALSE}
setwd("~/Data/Downloaded Data Sets/CPTACandKIRC_111523")

#First read in manifest that contains all file names.

Manifest <- read_delim("gdc_sample_sheet.2023-11-15.tsv") %>%
  select(1,2,5:8) %>%
  rename("FileID" = `File ID`,
         "Filename" = `File Name`,
         "Project" = `Project ID`,
         "Case" = `Case ID`,
         "Sample" = `Sample ID`,
         "Type" = `Sample Type`)

# Define groups of files for each project and category

CPTAC_N <- Manifest %>%
  filter(Project == "CPTAC-3", Type == "Solid Tissue Normal")

CPTAC_T <- Manifest %>%
  filter(Project == "CPTAC-3", Type == "Primary Tumor")

KIRC_N <- Manifest %>%
  filter(Project == "TCGA-KIRC", Type == "Solid Tissue Normal")

KIRC_T <- Manifest %>%
  filter(Project == "TCGA-KIRC", Type == "Primary Tumor")

```

This analysis includes `r nrow(CPTAC_N)` normal samples and `r nrow(CPTAC_T)` primary tumor samples from the CPTAC-3 project, and `r nrow(KIRC_N)` normal kidney samples and `r nrow(KIRC_T)` primary tumor samples from the TCGA KIRC project.

```{r read in data from CPTAC normal kidneys, include=FALSE}

GeneList <- c("ARNT", "ARNT2", "ARNTL", "ARNTL2",
              "VEGFA", "NR1D1", "PER2", "DKELR3", "DNTA", "SLC2A1",
              "SERPINE1")

folder <- CPTAC_N$FileID[1]
file <- CPTAC_N$Filename[1]
f_path <- str_glue("gdc_download_20231116_002116.005776/{folder}/{file}")

Data_CPTAC_N <- read_delim(f_path,
                           skip = 6,
                           col_names = c(
                             "gene_id",
                             "gene_name",
                             "gene_type",
                             "unstranded",
                             "stranded_first",
                             "stranded_second",
                             "tpm_unstranded",
                             "fpkm_unstranded",
                             "fpkm_uq_unstranded")
                           ) %>%
  filter(gene_name %in% GeneList)

Data_CPTAC_N$source_file <- file
  
N <- nrow(CPTAC_N)
  for(i in 2:N){
    folder <- CPTAC_N$FileID[i]
    file <- CPTAC_N$Filename[i]
    f_path <- str_glue("gdc_download_20231116_002116.005776/{folder}/{file}")
    Data_Temp <- read_delim(f_path,
                           skip = 6,
                           col_names = c(
                             "gene_id",
                             "gene_name",
                             "gene_type",
                             "unstranded",
                             "stranded_first",
                             "stranded_second",
                             "tpm_unstranded",
                             "fpkm_unstranded",
                             "fpkm_uq_unstranded")
                           ) %>%
  filter(gene_name %in% GeneList)
    
    Data_Temp$source_file <- file

    Data_CPTAC_N <- Data_CPTAC_N %>%
      rbind(Data_Temp)
    }

Data_CPTAC_N_Sel <- Data_CPTAC_N %>% select(2,7)
Data_CPTAC_N_Sel$Status <- "Normal"

```


```{r read in data from CPTAC tumors, include=FALSE}

folder <- CPTAC_T$FileID[1]
file <- CPTAC_T$Filename[1]
f_path <- str_glue("gdc_download_20231116_002116.005776/{folder}/{file}")

Data_CPTAC_T <- read_delim(f_path,
                           skip = 6,
                           col_names = c(
                             "gene_id",
                             "gene_name",
                             "gene_type",
                             "unstranded",
                             "stranded_first",
                             "stranded_second",
                             "tpm_unstranded",
                             "fpkm_unstranded",
                             "fpkm_uq_unstranded")
                           ) %>%
  filter(gene_name %in% GeneList)

Data_CPTAC_T$source_file <- file
  
N <- nrow(CPTAC_T)
  for(i in 2:N){
    folder <- CPTAC_T$FileID[i]
    file <- CPTAC_T$Filename[i]
    f_path <- str_glue("gdc_download_20231116_002116.005776/{folder}/{file}")
    Data_Temp <- read_delim(f_path,
                           skip = 6,
                           col_names = c(
                             "gene_id",
                             "gene_name",
                             "gene_type",
                             "unstranded",
                             "stranded_first",
                             "stranded_second",
                             "tpm_unstranded",
                             "fpkm_unstranded",
                             "fpkm_uq_unstranded")
                           ) %>%
  filter(gene_name %in% GeneList)
    
    Data_Temp$source_file <- file

    Data_CPTAC_T <- Data_CPTAC_T %>%
      rbind(Data_Temp)
    }

Data_CPTAC_T_Sel <- Data_CPTAC_T %>% select(2,7)
Data_CPTAC_T_Sel$Status <- "Tumor"

```

```{r combine normal and tumor tibbles for CPTAC, include=FALSE}

List_Data_CPTAC3 <- rbind(Data_CPTAC_N_Sel, Data_CPTAC_T_Sel) %>%
  rename("TPM" = tpm_unstranded)

```

```{r run t-test and create boxplot for CPTAC, include=FALSE}

stat.test <- List_Data_CPTAC3 %>% group_by(gene_name) %>%
    t_test(TPM ~ Status) %>%
    add_significance()
stat.test

p1 <- ggplot(List_Data_CPTAC3, aes(gene_name, TPM, fill = Status))

```

Results for CPTAC-3:

```{r embed results for CPTAC-3}
p1 + 
  geom_boxplot() + 
  scale_fill_manual(values = c("black", "red")) +
  labs(title = "CPTAC-3") +
  xlab("") +
  theme_classic() + 
  theme(axis.text.x = element_text(face = "italic",
                                size = 10))

stat.test
```


```{r read in data from KIRC normal kidneys, include=FALSE}

folder <- KIRC_N$FileID[1]
file <- KIRC_N$Filename[1]
f_path <- str_glue("gdc_download_20231116_002116.005776/{folder}/{file}")

Data_KIRC_N <- read_delim(f_path,
                           skip = 6,
                           col_names = c(
                             "gene_id",
                             "gene_name",
                             "gene_type",
                             "unstranded",
                             "stranded_first",
                             "stranded_second",
                             "tpm_unstranded",
                             "fpkm_unstranded",
                             "fpkm_uq_unstranded")
                           ) %>%
  filter(gene_name %in% GeneList)

Data_KIRC_N$source_file <- file
  
N <- nrow(KIRC_N)
  for(i in 2:N){
    folder <- KIRC_N$FileID[i]
    file <- KIRC_N$Filename[i]
    f_path <- str_glue("gdc_download_20231116_002116.005776/{folder}/{file}")
    Data_Temp <- read_delim(f_path,
                           skip = 6,
                           col_names = c(
                             "gene_id",
                             "gene_name",
                             "gene_type",
                             "unstranded",
                             "stranded_first",
                             "stranded_second",
                             "tpm_unstranded",
                             "fpkm_unstranded",
                             "fpkm_uq_unstranded")
                           ) %>%
  filter(gene_name %in% GeneList)
    
    Data_Temp$source_file <- file

    Data_KIRC_N <- Data_KIRC_N %>%
      rbind(Data_Temp)
    }

Data_KIRC_N_Sel <- Data_KIRC_N %>% select(2,7)
Data_KIRC_N_Sel$Status <- "Normal"

```


```{r read in data from KIRC tumors, include=FALSE}

folder <- KIRC_T$FileID[1]
file <- KIRC_T$Filename[1]
f_path <- str_glue("gdc_download_20231116_002116.005776/{folder}/{file}")

Data_KIRC_T <- read_delim(f_path,
                           skip = 6,
                           col_names = c(
                             "gene_id",
                             "gene_name",
                             "gene_type",
                             "unstranded",
                             "stranded_first",
                             "stranded_second",
                             "tpm_unstranded",
                             "fpkm_unstranded",
                             "fpkm_uq_unstranded")
                           ) %>%
  filter(gene_name %in% GeneList)

Data_KIRC_T$source_file <- file
  
N <- nrow(KIRC_T)
  for(i in 2:N){
    folder <- KIRC_T$FileID[i]
    file <- KIRC_T$Filename[i]
    f_path <- str_glue("gdc_download_20231116_002116.005776/{folder}/{file}")
    Data_Temp <- read_delim(f_path,
                           skip = 6,
                           col_names = c(
                             "gene_id",
                             "gene_name",
                             "gene_type",
                             "unstranded",
                             "stranded_first",
                             "stranded_second",
                             "tpm_unstranded",
                             "fpkm_unstranded",
                             "fpkm_uq_unstranded")
                           ) %>%
  filter(gene_name %in% GeneList)
    
    Data_Temp$source_file <- file

    Data_KIRC_T <- Data_KIRC_T %>%
      rbind(Data_Temp)
    }

Data_KIRC_T_Sel <- Data_KIRC_T %>% select(2,7)
Data_KIRC_T_Sel$Status <- "Tumor"

```

```{r combine normal and tumor tibbles for KIRC, include=FALSE}

List_Data_KIRC <- rbind(Data_KIRC_N_Sel, Data_KIRC_T_Sel) %>%
  rename("TPM" = tpm_unstranded)

```

```{r run t-test and create boxplot for KIRC, include=FALSE}

stat.test <- List_Data_KIRC %>% group_by(gene_name) %>%
    t_test(TPM ~ Status) %>%
    add_significance()

p2 <- ggplot(List_Data_KIRC, aes(gene_name, TPM, fill = Status))

```

Results for TCGA-KIRC:

```{r embed results for TCGA-KIRC}
p2 + 
  geom_boxplot() + 
  scale_fill_manual(values = c("black", "red")) +
  labs(title = "TCGA-KIRC") +
  xlab("") +
  theme_classic() + 
  theme(axis.text.x = element_text(face = "italic",
                                size = 10))

stat.test

```


```{r combine all expression data, include=FALSE}

List_Data_CPTAC3$Project <- "CPTAC-3"
List_Data_KIRC$Project <- "TCGA_KIRC"

List_Data_combined <- rbind(List_Data_CPTAC3, List_Data_KIRC)

ARNT_Combined <- List_Data_combined %>%
  filter(gene_name == "ARNT")
ARNT2_Combined <- List_Data_combined %>%
  filter(gene_name == "ARNT2")
BMAL1_Combined <- List_Data_combined %>%
  filter(gene_name == "ARNTL")
BMAL2_Combined <- List_Data_combined %>%
  filter(gene_name == "ARNTL2")
REVERBa_Combined <- List_Data_combined %>%
  filter(gene_name == "NR1D1")
VEGFA_Combined <- List_Data_combined %>%
  filter(gene_name == "VEGFA")
PER2_Combined <- List_Data_combined %>%
  filter(gene_name == "PER2")

```

```{r run t-tests and create boxplots}

#ARNT
stat.test1 <- ARNT_Combined %>% group_by(Project) %>%
    t_test(TPM ~ Status) %>%
    add_significance()

bxp1 <- ggboxplot(
    ARNT_Combined, x = "Project", y = "TPM", 
    fill = "Status", palette =c("black", "red"),
    width = 0.7,
    ggtheme = theme_classic(),
    xlab = ""
)

# Add p-value and significance levels
stat.test1 <- stat.test1 %>% add_xy_position(x = "Project")
bxp1 + 
    stat_pvalue_manual(stat.test1, size = 4, tip.length = 0) +
    labs(y = expression(~italic(ARNT)~" (TPM)"))


#ARNT2
stat.test2 <- ARNT2_Combined %>% group_by(Project) %>%
    t_test(TPM ~ Status) %>%
    add_significance()

bxp2 <- ggboxplot(
    ARNT2_Combined, x = "Project", y = "TPM", 
    fill = "Status", palette =c("black", "red"),
    xlab = ""
)

# Add p-value and significance levels
stat.test2 <- stat.test2 %>% add_xy_position(x = "Project")
bxp2 + 
    stat_pvalue_manual(stat.test2, size = 8, tip.length = 0) +
    labs(y = expression(~italic(ARNT2)~" (TPM)"))

#BMAL1
stat.test3 <- BMAL1_Combined %>% group_by(Project) %>%
    t_test(TPM ~ Status) %>%
    add_significance()

bxp3 <- ggboxplot(
    BMAL1_Combined, x = "Project", y = "TPM", 
    fill = "Status", palette =c("black", "red"),
    xlab = ""
)

# Add p-value and significance levels
stat.test3 <- stat.test3 %>% add_xy_position(x = "Project")
bxp3 + 
    stat_pvalue_manual(stat.test3, size = 8, tip.length = 0) +
    labs(y = expression(~italic(BMAL1)~" (TPM)"))

#BMAL2
stat.test4 <- BMAL2_Combined %>% group_by(Project) %>%
    t_test(TPM ~ Status) %>%
    add_significance()

bxp4 <- ggboxplot(
    BMAL2_Combined, x = "Project", y = "TPM",
    fill = "Status", palette =c("black", "red"),
    xlab = ""
)

# Add p-value and significance levels
stat.test4 <- stat.test4 %>% add_xy_position(x = "Project")
bxp4 + 
    stat_pvalue_manual(stat.test4, size = 8, tip.length = 0) +
    labs(y = expression(~italic(BMAL2)~" (TPM)"))




#REVERBa
stat.test5 <- REVERBa_Combined %>% group_by(Project) %>%
    t_test(TPM ~ Status) %>%
    add_significance()

bxp5 <- ggboxplot(
    REVERBa_Combined, x = "Project", y = "TPM",
    fill = "Status", palette =c("black", "red"),
    xlab = ""
)

# Add p-value and significance levels
stat.test5 <- stat.test5 %>% add_xy_position(x = "Project")
bxp5 + 
    stat_pvalue_manual(stat.test5, size = 8, tip.length = 0) +
    labs(y = expression(~italic(REVERBa)~" (TPM)"))

#VEGFA
stat.test6 <- VEGFA_Combined %>% group_by(Project) %>%
    t_test(TPM ~ Status) %>%
    add_significance()

bxp6 <- ggboxplot(
    VEGFA_Combined, x = "Project", y = "TPM",
    fill = "Status", palette =c("black", "red"),
    xlab = ""
)

# Add p-value and significance levels
stat.test6 <- stat.test6 %>% add_xy_position(x = "Project")
bxp6 + 
    stat_pvalue_manual(stat.test6, size = 8, tip.length = 0) +
    labs(y = expression(~italic(VEGFA)~" (TPM)"))

#PER22
stat.test7 <- PER2_Combined %>% group_by(Project) %>%
    t_test(TPM ~ Status) %>%
    add_significance()

bxp7 <- ggboxplot(
    PER2_Combined, x = "Project", y = "TPM",
    fill = "Status", palette =c("black", "red"),
    xlab = ""
)

# Add p-value and significance levels
stat.test7 <- stat.test7 %>% add_xy_position(x = "Project")
bxp7 + 
    stat_pvalue_manual(stat.test7, size = 8, tip.length = 0) +
    labs(y = expression(~italic(PER2)~" (TPM)"))

```

